Learning Fast Quadruped Robot Gaits with the RL PoWER Spline Parameterization

### 2012-06-15 paper_aimsa/abstract.tex
submission to Petar

Legged robots are uniquely privileged over their wheeled counterparts
in their potential to access rugged terrain. However, designing
walking gaits by hand for legged robots is a difficult and
time-consuming process, so we seek algorithms for learning such gaits
automatically using real world experimentation. Numerous previous
studies have examined a variety algorithms for learning gaits, using
an assortment of different robots. It is often difficult to compare
the algorithmic results from one study to the next, because the
conditions and robots used vary. With this in mind, we have used an
open-source, 3D printed quadruped robot called QuadraTot, so the results may be
verified, and hopefully improved upon, by any group so
desiring. Because many robots do not have accurate simulators, we test
gait-learning algorithms entirely on the physical robot. Previous
studies using the QuadraTot have compared parameterized splines, the
HyperNEAT generative encoding, and (Kyrre's algorithm). Here we
compare these results to an algorithm called Policy learning by
Weighting Exploration with the Returns, or RL PoWER. We report that
this algorithm has learned the fastest gait yet reported in the
literature, 16.3\% faster \edit{different number than in results?} than reported for HyperNEAT and \edit{???
compared to Kyrre's?}. In addition, the learned gaits are less taxing
on the robot and more repeatable than previous record-breaking gaits.

>>>>

Legged robots are uniquely privileged over their wheeled counterparts
in their potential to access rugged terrain. However, designing
walking gaits by hand for legged robots is a difficult and
time-consuming process, so we seek algorithms for learning such gaits
automatically using real world experimentation. Numerous previous
studies have examined a variety algorithms for learning gaits, using
an assortment of different robots. It is often difficult to compare
the algorithmic results from one study to the next, because the
conditions and robots used vary. With this in mind, we have used an
open-source, 3D printed quadruped robot called QuadraTot, so the results may be
verified, and hopefully improved upon, by any group so
desiring. Because many robots do not have accurate simulators, we test
gait-learning algorithms entirely on the physical robot. Previous
studies using the QuadraTot have compared parameterized splines, the
HyperNEAT generative encoding and genetic algorithm. Among these, the research on the genetic algorithm was conducted by \citep{glette2012evolution-of-locomotion-in-a-simulated} in a simulator and tested on a real robot. Here we
compare these results to an algorithm called Policy learning by
Weighting Exploration with the Returns, or RL PoWER. We report that
this algorithm has learned the fastest gait through phsysical experiments yet reported in the
literature, 16.3\% faster than reported for HyperNEAT. In addition, the learned gaits are less taxing
on the robot and more repeatable than previous record-breaking gaits.

### 2012-06-15 paper_aimsa/conclusion.tex
submission to Petar

We have presented results from a recently introduced reinforcement-learning-based algorithm for
optimizing a quadrupedal gait for linear speed, tested on a physical robot. \edit{Next sentence mistakenly copied?} We implemented and
tested six learning strategies for parameterized gaits and compared
them to gaits produced by neural networks evolved with the HyperNEAT
generative encoding. Though both methods resulted in an improvement
over the robot’s previous \naive gaits, based on the statistics
collected, RL PoWER has a more elegant and consistent performance.

Over 900 trials have been made to investigate the applicability
of RL PoWER to quadruped robots. It is difficult to gather the enough
trials that would be necessary to properly rank the methods. One
direction for future work could be to obtain many more trials. But due
to the physical limitations, obtaining one solution to this is
simulation. Because of the low cost of simulation, it would produce
the necessary volume of trials to allow the learning methods to be
effective, and the hardware trials would serve to continuously ground
and refine the simulator. \cite{glette2012evolution-of-locomotion-in-a-simulated} \edit{what is this citation for?}

One view supported by this study is that for feedback oriented tasks,
reinforcement learning methods are more fit and natural. Despite the
complexities of HyperNEAT, a simpler algorithm on the code level \edit{???}
delivered better performance in general.  Also, evolvable spline
interpolation is shown to be simple and representationally powerful at
the same time. Evolvable splines can serve as a general representation for various other
learning problems.\edit{probably rewrite this last paragraph}

>>>>

We have presented results from a recently introduced reinforcement-learning-based algorithm for
optimizing a quadrupedal gait for linear speed, tested on a physical robot.  We implemented the algorithm, Policy learning by Weighting
Exploration with the Returns (a.k.a RL PoWER), for parameterized gaits and compared
it to gaits produced by neural networks evolved with the HyperNEAT
generative encoding. Though both methods resulted in an improvement
over the robot’s previous \naive gaits, based on the statistics
collected, RL PoWER has a more elegant and consistent performance.

Over 900 trials have been made to investigate the applicability
of RL PoWER to quadruped robots. It is difficult to gather the enough
trials that would be necessary to properly rank the methods. One
direction for future work could be to obtain many more trials. But due
to the physical limitations, obtaining one solution to this is
simulation. Because of the low cost of simulation, it would produce
the necessary volume of trials to allow the learning methods to be
effective, and the hardware trials would serve to continuously ground
and refine the simulator.

One view supported by this study is that for feedback oriented tasks,
reinforcement learning methods are more fit and natural. Despite the
complexities of HyperNEAT, a structurally simpler algorithm delivered better performance in general.  Also, evolvable spline
interpolation is shown to be simple and representationally powerful at
the same time. Evolvable splines can serve as a general representation for various other
learning problems.

### 2012-06-15 paper_aimsa/experimentalSetup.tex
submission to Petar

The quadruped robot used, QuadraTot, was assembled from parts purchased online and
parts printed by the Objet Connex 500 3-D Printing System. The robot
actuation system consists of 5 AX-18+ Dynamixel servos and 4 AX-12+
Dynamixel servos: one inner joint with one AX-18+ servo and one outer
joint with AX-12+servo in each of the four legs, and one AX-18+ servo
at the center. To avoid the formerly reported \edit{???} AX-18+ servos are used
in this robot because of their stronger actuation power than that of
AX-12+. Each servo could be set to a position in the range [0, 1023]
by using pydynamixel library, corresponding roughly to a physical range [-120$^\circ$,
  +120$^\circ$]. Also, to prevent collisions with the robot body, the control
module filter out the commands to a safe range. This range was [-85$^\circ$,
  +60$^\circ$] for the inner leg servos, [-113$^\circ$, +39$^\circ$] for the outer leg
servos, and [-28$^\circ$, +28$^\circ$] for the center hip servo. In the studies of
this paper, tethered cables powered both the computer and the
servos. It measures approximately 39.5 centimeters from leg to
opposite leg in the crouch position showed in \figref{quadratotWhiteBooties}.

Our performance metric was the displacement over the evaluation period of 12 seconds
for each. Same as \cite{yosinski2011evolving-robot-gaits}, the displacement was
measured using a Wii remote that was placed on the ceiling. Different
from the original model described in \cite{yosinski2011evolving-robot-gaits}, the quadruped robot was
equipped with a three-infrared-LED cluster on top rather than just
one. The reason for this setup is that when fierce gaits were
executed, the Wii remote loses tracking of the robot position due to
the limited visible angle of a single LED.  These three LEDs were
place tightly together to act as one signal emitter. Each LED was
tilted outwards in order to maximize the visible range. A
separate tracking server ran on the robot PC interacted with the Wii
remote via bluetooth by using the CWiid library.  A corresponding
client communicates with this server via socket connection. Our
fitness evaluation code talks with the Wii remote by using this client
and updates the position from beginning to end during each run. The
metric for evaluating gaits was the Euclidian distance the robot
traveled during a 12-second run on flat surface. Previous work done in
\cite{yosinski2011evolving-robot-gaits} and \cite{clune2009evolving-coordinated-quadruped} suggested that extremely fierce gaits are not viable in
general. These gaits tend to either overburden the servos or flip the
robot. \edit{Explain this next part on smart cropping better.} Thus, RL PoWER were tested after carefully cropping out the
extremely fierce gaits each time. As described earlier, each leg has
two joints, inner joint j1 and outer j2. The position of one leg is
determined by the sum of the values of j1 and j2. Extremely fierce
gaits usually have relatively small values of $g$. As shown in \figref{smartCrop}, this cropping method works
by mapping the wild gaits onto the boundary of a quasi-triangular area in
the two-dimensional space of j1 and j2.

\figp{smartCrop}{1}{The red line is the cropping border, $g = j1 +  j2$. As shown in \figref{sixLegPositions} \edit{???}}

\figp{sixLegPositions}{1}{The left two positions are allowed; middle
  two positions are on the red border line from \figref{smartCrop};
  right two positions are deemed as ``extremely fierce'', thus are
  disallowed.   \\\editbox{Smart crop needs to be explained somewhere, probably briefly in this caption and more extensively in the text.}}

>>>>

The quadruped robot used, QuadraTot, was assembled from parts purchased online and
parts printed by the Objet Connex 500 3-D Printing System. The robot
actuation system consists of 5 AX-18+ Dynamixel servos and 4 AX-12+
Dynamixel servos: one inner joint with one AX-18+ servo and one outer
joint with AX-12+servo in each of the four legs, and one AX-18+ servo
at the center. To avoid the formerly reported problem with  AX-18+ servos are used
in this robot because of their stronger actuation power than that of
AX-12+. Each servo could be set to a position in the range [0, 1023]
by using pydynamixel library, corresponding roughly to a physical range [-120$^\circ$,
  +120$^\circ$]. Also, to prevent collisions with the robot body, the control
module filter out the commands to a safe range. This range was [-85$^\circ$,
  +60$^\circ$] for the inner leg servos, [-113$^\circ$, +39$^\circ$] for the outer leg
servos, and [-28$^\circ$, +28$^\circ$] for the center hip servo. In the studies of
this paper, tethered cables powered both the computer and the
servos. It measures approximately 39.5 centimeters from leg to
opposite leg in the crouch position showed in \figref{quadratotWhiteBooties}.

Our performance metric was the displacement over the evaluation period of 12 seconds
for each. Same as \cite{yosinski2011evolving-robot-gaits}, the displacement was
measured using a Wii remote that was placed on the ceiling. Different
from the original model described in \cite{yosinski2011evolving-robot-gaits}, the quadruped robot was
equipped with a three-infrared-LED cluster on top rather than just
one. The reason for this setup is that when fierce gaits were
executed, the Wii remote loses tracking of the robot position due to
the limited visible angle of a single LED.  These three LEDs were
place tightly together to act as one signal emitter. Each LED was
tilted outwards in order to maximize the visible range. A
separate tracking server ran on the robot PC interacted with the Wii
remote via bluetooth by using the CWiid library.  A corresponding
client communicates with this server via socket connection. Our
fitness evaluation code talks with the Wii remote by using this client
and updates the position from beginning to end during each run. The
metric for evaluating gaits was the Euclidian distance the robot
traveled during a 12-second run on flat surface. Previous work done in
\cite{yosinski2011evolving-robot-gaits} and \cite{clune2009evolving-coordinated-quadruped} suggested that extremely fierce gaits are not viable in
general. These gaits tend to either overburden the servos or flip the
robot. Thus, RL PoWER were tested after carefully cropping out the
extremely fierce gaits each time. As described earlier, each leg has
two joints, inner joint $j1$ and outer $j2$. The position of one leg is
determined by the sum of the values of $j1$ and $j2$, in the range of [0, 1023]. Extremely fierce
gaits usually have relatively small values of $g$. As shown in \figref{smartCrop}, this cropping method works
by mapping the wild gaits right onto the boundary of a quasi-triangular area in
the two-dimensional space of $j1$ and $j2$. In our experimentation, we set the threshhold to 730 to balance between gait performance and safety. However, an obvious trade-off to this method is the reduced exploratory region of the parameters. A extension would be experiments using smaller $g$, cropping less gaits, to possibly enhance the performance.

\figp{smartCrop}{0.7}{The red line is the smart cropping border, $g = j1 +  j2$. The dashed grey lines are the physical limits of the joints. As shown in \figref{sixLegPositions}, the fierce movements are prevented by mapping these out-of-range positions to the border line as shown in the graph. The value of $g$ cannot drop below 730 otherwise $j1$ and $j2$ are projected right to the red line.  }

\figp{sixLegPositions}{1}{The left two positions are allowed; middle
  two positions are on the red border line from \figref{smartCrop};
  right two positions are deemed as ''extremely fierce'', thus are
  disallowed. Future work may involve studying a larger range of $g$ for increased performance.}

### 2012-06-15 paper_aimsa/main.tex
submission to Petar

%\title{RL-PoWER, Evolving Policy Parameterization with Reinforcement Learning and its Application to Quadrupedal Walking Optimization}
\title{Learning Fast Quadruped Robot Gaits with the RL PoWER Spline Parameterization}
\author{Haocheng Shen, Jason Yosinski, Petar Kormushev, \and Hod Lipson  \\
  \mbox{}\\
Cornell University, 239 Upson Hall, Ithaca, NY 14853 \\
\texttt{hs454@cornell.edu,yosinski@cs.cornell.edu}}

>>>>

%\title{RL-PoWER, Evolving Policy Parameterization with Reinforcement Learning and its Application to Quadrupedal Walking Optimization}
\title{Learning Fast Quadruped Robot Gaits with the RL PoWER Spline Parameterization}
\author{Haocheng Shen, Jason Yosinski, Petar Kormushev, Darwin G. Caldwell \and Hod Lipson  \\
  \mbox{}\\
Cornell University, 239 Upson Hall, Ithaca, NY 14853, USA\\
Department of Advanced Robotics, Istituto Italiano di Tecnologia, via Morego 30, 16163 Genova, Italy \\
\texttt{hs454@cornell.edu, Petar.Kormushev@iit.it}}

### 2012-06-15 paper_aimsa/methods.tex
submission to Petar

\subsection{Policy Representation by Splines}
The simplest model with back-compatibility is geometric
splines. \editbox{Too fast... add a sentence or two on what the methods are trying to do, and what back-compatibility is.} For a given model f(x) with K knots, we can preserve the
exact shape of the generated curve while adding extra knots to the
original spline. Say, if we put one additional knot between every two
consecutive knots of the original spline, we end up with a 2K - 1
knots and a spline that has the same shape as the original one. In
order to do this, we need to define an algorithm for evolving the
parameterization from $K$ to $L$ knots ($L > K$), which is formulated as 
Algorithm 1 in \cite{kormushev2011bipedal-walking-energy} \edit{reproduce that algorithm here?}.  Without loss of generality, the policy parameters are
normalized into $[0, 1]$, and appropriately scaled/shifted as necessary
later upon use.

@@

\cite{kober2009learning-motor-primitives} proposed a RL algorithm
called Policy learning by Weighting Exploration with the
Returns (RL PoWER), which is based on Expectation-Maximization algorithm
(EM). The proposed technique for evolving the policy parameterization
is a combination with this EM-based RL algorithm, named PoWER \cite{kober2009learning-motor-primitives}. \edit{Check the prev sentence... I don't get it} The reason for using this is its relatively fewer parameters that need tuning. We
evolved the policy parameterization only on those past trials ranked
the highest by the importance sampling technique used by the PoWER
algorithm. The intuition behind is that highly ranked
parameterizations have more potential to evolve even better in the
future. Besides, evolving all the parameterizations increases the exploring space. Since our experiment
is done on a physical robot, exploring all the variations of every
parameterization is not practical. Future work may incorporate simulations into the studies, as illustrated in \cite{bongard2006resilient-machines-through}.

For the experiment, we set the splines to have 3 knots for each servo, and there are 8
servos in total. The servo in the hip is not used in
our experiment. Previous work has verified that quadruped gaits
perform better when they are coordinated \citep{clune2009evolving-coordinated-quadruped, clune2011on-the-performance-of-indirect-encoding, valsalam2008modular-neuroevolution-for-multilegged}. For each spline, we calculate its corresponding parameterized gait for one unit time cycle \edit{how long is this?} and then apply the same
pattern to every cycle throughout the 12 seconds of one
trial. Specifically, each spline (a set of 3 knots) is interpreted to its corresponding servo positions as shown in \eqnref{rl} and \tabref{params}.

@@

\be
\vec{g}(t) =
\left[ {\begin{array}{c@{ }c@{ }c@{ }l@{ }l}
R \cdot f(s1, s2, s3) & \ \          & \             & \            & + C \\ % 1
0                              & \             & \             & \            & + C_C \\ % 8
\end{array} } \right]
\eqnlabel{rl}
\ee

\editbox{Add a new equation showing all servos (all nine) with their
  actual values of the parameters $R$ and $C$ plugged in.}

>>>>

\subsection{Policy Representation by Splines}
The simplest model with back-compatibility is geometric
splines. For example, for a given model f(x) with K knots, we can preserve the
exact shape of the generated curve while adding extra knots to the
original spline. If we put one additional knot between every two
consecutive knots of the original spline, we end up with a 2K - 1
knots and a spline that has the same shape as the original one. In
order to do this, we need to define an algorithm for evolving the
parameterization from $K$ to $L$ knots ($L > K$), which is formulated as in
\ref{alg_Evolve_Spline}.  Without loss of generality, the policy parameters are
normalized into $[0, 1]$, and appropriately scaled and shifted as necessary upon use.

\begin{algorithm}
\caption{EvolvePolicy-Spline ($P_{current}$: current policy, $L$: desired new number of parameters)}
\label{alg_Evolve_Spline}
\begin{algorithmic}[1]

\STATE $K$ $\leftarrow$ $P_{current}.numberOfParameters$

\STATE $X_{current}$ $\leftarrow$ $[0, \frac{1}{K-1}, \frac{2}{K-1}, ... , 1]$ % Knots
\STATE $Y_{current}$ $\leftarrow$ $P_{current}.parameterValues$ % Values
\STATE $S_{current}$ $\leftarrow$ ConstructSpline($X_{current}$, $Y_{current}$) % Spline
%\COMMENT{ Optional: make the spline cyclic}

\STATE $X_{new}$ $\leftarrow$ $[0, \frac{1}{L-1}, \frac{2}{L-1}, ... , 1]$
\STATE $Y_{new}$ $\leftarrow$ EvaluateSplineAtKnots($S_{current}$, $X_{new}$)
\STATE $S_{new}$ $\leftarrow$ ConstructSpline($X_{new}$, $Y_{new}$)
%\COMMENT{ Optional: make the spline cyclic}

\STATE $P_{new}.numberOfParameters$ $\leftarrow$ $L$
\STATE $P_{new}.parameterValues$ $\leftarrow$ $S_{new}$.$Y_{new}$

\STATE return $P_{new}$

\end{algorithmic}
\end{algorithm}

@@

\cite{kober2009learning-motor-primitives} proposed a RL algorithm
named Policy learning by Weighting Exploration with the
Returns (RL PoWER), which is based on Expectation-Maximization algorithm
(EM). The reason for using this is its relatively fewer parameters that need tuning. We
evolved the policy parameterization only on those past trials ranked
the highest by the importance sampling technique used by the PoWER
algorithm. The intuition behind is that highly ranked
parameterizations have more potential to evolve even better in the
future. Besides, evolving all the parameterizations increases the exploring space. Since our experiment
is done on a physical robot, exploring all the variations of every
parameterization is not practical. Future work may incorporate simulations into the studies, as illustrated in \cite{bongard2006resilient-machines-through}.

For the experiment, we set the splines to have 3 knots for each servo, and there are 8
servos in total. The servo in the hip is not used in
our experiment. Previous work has verified that quadruped gaits
perform better when they are coordinated \citep{clune2009evolving-coordinated-quadruped, clune2011on-the-performance-of-indirect-encoding, valsalam2008modular-neuroevolution-for-multilegged}. For each spline, we calculate its corresponding parameterized gait for one unit time cycle of 1.8 seconds and then apply the same
pattern to every cycle throughout the 12 seconds of one
trial. Specifically, each spline (a set of 3 knots) is interpreted to its corresponding servo positions as in the following equation and in \tabref{params}.

@@

\[
\vec{g}(t) =
\left[ {\begin{array}{c@{ }c@{ }c@{ }l@{ }l}
R \cdot f(s1, s2, s3) & \ \          & \             & \            & + C \\ % 1
0                              & \             & \             & \            & + C_C \\ % 8
\end{array} } \right]
\]

### 2012-06-15 paper_aimsa/relatedWork.tex
submission to Petar

Many attempts have been made on robot gaits learning using machine
learning algorithms, producing competitive results \edit{Change these
  all to citations} (Chernova and Veloso, 2005; Hornby et al., 2005;
Zykov et al., 2004; Clune et al., 2009a, 2011, 2009b,c; T´ellez et
al., 2006; Valsalam and Miikkulainen, 2008). In fact, Jason et
al. tested and compared six different algorithms for quadruped robot
gait learning in their studies. HyperNEAT, a generative encoding
neural network algorithm, achieved the best performance among all six
methods. In Kormushev et al. (2011)'s bipedal robot energy reduction
research, a reinforcement learning method was used to optimize the
performance. The results from Kormushev's research showed this
algorithm's potential for walking problems. This reinforcement
learning method still needs more tests for further explorations.

>>>>

Many attempts have been made on robot gaits learning using machine
learning algorithms, producing competitive results (Chernova and Veloso, 2005; Hornby et al., 2005;
Zykov et al., 2004; Clune et al., 2009a, 2011, 2009b,c; T´ellez et
al., 2006; Valsalam and Miikkulainen, 2008). In fact, Jason et
al. tested and compared six different algorithms for quadruped robot
gait learning in their studies. HyperNEAT, a generative encoding
neural network algorithm, achieved the best performance among all six
methods. In Kormushev et al. (2011)'s bipedal robot energy reduction
research, a reinforcement learning method was used to optimize the
performance. The results from Kormushev's research showed this
algorithm's potential for walking problems. This reinforcement
learning method still needs more tests for further explorations.

### 2012-06-15 paper_aimsa/results.tex
submission to Petar

\begin{table}
\begin{center}
\begin{tabular}{|r|c|c|c||c|}
\hline
                                         & Average & Std. Dev \\
\hline                                    
\hline                                    
RL PoWER                       & {\bf 7.62}  &    2.1   \\
\hline
HyperNEAT                      & 6.28   &   1.26   \\
\hline
\end{tabular}
\caption{The average and standard deviation of the best gaits found
  for both algorithm during each of three runs, in cm/sec. \edit{Add the best gait each algorithm found as well.}}
\tablabel{results}
\end{center}
\end{table}

We believe that this is because RL PoWER uses evolvable splines to represent gaits As explained earlier,  
The single best gait found during this study had a speed of 11.09 centimeters/sec,  14.7\% better \edit{different number than in abstract?} than the best HyperNEAT gait. \figref{rlPattern} shows a typical RL PoWER gait that had high fitness. Compared with \figref{neat_110115_211410_00000_002_filt_zoom}, the pattern of RL PoWER's motion is less
complex but more regular than the HyperNEAT, but producing higher speed. As shown in \figref{rlPattern}
, multiple motors are more coordinated than the HyperNEAT gait.

\figp{rlPattern}{0.75}{The motor positions of a typical well
  performing gait over one cycle. It matches exactly the shape of its
  corresponding spline representation \edit{what does this sentence mean?}. This pattern shows the
  simplicity and power of geometric splines representations \edit{how does it show this?}.}

>>>>

\begin{table}
\begin{center}
\begin{tabular}{|r|c|c|c||c|}
\hline
                                & Average & Best Single & Std. Dev \\
\hline                                    
\hline                                    
RL PoWER                       & {\bf 7.62}  & 11.05  &    2.1   \\
\hline
HyperNEAT                      & 6.28   & 9.7 & 1.26   \\
\hline
\end{tabular}
\caption{The average and standard deviation of the best gaits found
  for both algorithm during each of three runs, in cm/sec.}
\tablabel{results}
\end{center}
\end{table}

We believe that this is because RL PoWER uses evolvable splines to represent gaits As explained earlier,  
The single best gait found during this study had a speed of 11.09 cm/sec,  16.3\% better than the best HyperNEAT gait. \figref{rlPattern} shows a typical RL PoWER gait that had high fitness. Compared with \figref{neat_110115_211410_00000_002_filt_zoom}, the pattern of RL PoWER's motion is less
complex but more regular than the HyperNEAT, but producing higher speed. As shown in \figref{rlPattern}
, multiple motors are more coordinated than the HyperNEAT gait.

\figp{rlPattern}{0.75}{The motor positions of a typical well
performing gait over one cycle. According to the spline transformation, the curves of joint positions are merely linear transformations of the spline curves. Compared with the pattern of a well-performing HyperNEAT gait, the spline representation of the parameterized gaits are simple yet powerful.}
